---
title: "Nested ANOVA and MANOVA"
author: Rui Zhou
documentclass: ctexart
output:
  rticles::ctex:
    fig_caption: yes
    number_sections: yes
    toc: yes
---
# Nested ANOVA
## 0

```{r message=FALSE}
library(tidyverse)
library(car)

nestedData = readxl::read_excel("nested anova.xlsx")
```

## visualization
```{r}
# we should eyeball the data first to make sense of the test.
library(ggplot2)
# create boxplots to visualize the data
ggplot(nestedData, aes(x=factor(location), y=DV)) + 
  geom_boxplot()
```

## nested ANOVA
Theoretically, the order of computing the variance sum of squares matters, thus appointing the `summation = 'I'`.

However, here the summation type does not matter.

Note that the result of the main effect of the group is wrong. We can manually make it straight from the SS of the group and the subgroup.
```{r}
# note that nested anova can use error type I and III.
# aov() has a default type I to process factors in an sequential order
# here type I and III don't differ, since effectively we have no orders
aov_model = aov(DV ~ factor(method)/factor(location), 
                data = nestedData,summation="I")
summary(aov_model)
aov_model = aov(DV ~ factor(method)/factor(location), 
                data = nestedData,summation="III")
summary(aov_model)
```

### error adjustment
An alternatDVe refinement is to adjust the error term.
```{r}
aov.result2 = aov(DV ~ factor(method) + 
                    Error(factor(method):factor(location)),
                  data = nestedData)
summary(aov.result2)
```

### comparison with one-way
If we wrongly ignore the nested factor and use one-way ANOVA, the results would be different

Note that there's only one group factor, therefore, the type of summation does not matter.
```{r}
aov_model = aov(DV ~ factor(method), 
                data = nestedData,summation="III")
summary(aov_model)

aov_model = aov(DV ~ factor(method), 
                data = nestedData)
summary(aov_model)
```

# MANOVA
## 0

```{r}
library("readxl")
library("rstatix")
library("MVTests")
library("heplots")

# salary data example
salaryData = readxl::read_xlsx("example.xlsx")
DV = cbind(salaryData$salary, salaryData$salbegin)
jobcat = as.factor(salaryData$jobcat)
gender = as.factor(salaryData$gender)
```

## visualization
Note that in order to generate the boxplot according to the "categorical" data, "factor" type should be designated in the mapping parameter.

If not "factored", the scatter plot works fine with grouped effect, but the boxplot fails.
```{r}
# eyeball the data
library(gridExtra)
p1 = ggplot(salaryData, aes(x = factor(jobcat), y = salary)) + 
  geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2) + 
  theme(legend.position="top")
p2 = ggplot(salaryData, aes(x = factor(jobcat), y = salbegin)) + 
  geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2) + 
  theme(legend.position="top")
grid.arrange(p1, p2, ncol=2)
```

We can also eyeball the DVs.
```{r}
library(car)
scatterplot(salary ~ salbegin, data=salaryData, ellipse=TRUE, 
  smooth=list(style="lines"))
```

## assumptions

### normality
#### pipe-friendly (Preferred)
Preferred!

Note that `mshapiro_test()` is from the package `rstatix`, one step further from `tidyverse`.
```{r}
salaryData |> select(salary,salbegin) |> mshapiro_test()

salaryData |> group_by(jobcat) |> 
  shapiro_test(salary,salbegin)
```

#### classical (Not Preferred)
Not preferred. It's not efficient to do this way.

The specific function `mshapiro.test()` requires to library the package `mvnormtest`, which surprisingly has a single function! Moreover, it requires the data to be transposed before being processed.

The most annoying part is that the data has to be packed purposefully.
```{r}
mvnormtest::mshapiro.test(t(DV))
```

### homogeniety of variance & covariance
Using the Box's M-test and Levene's Test.

#### pipe-friendly (Preferred)
Using `box_m()` from the package `rstatix` to test the homogeniety of covariance matrices, in which the first parameter is the DVs needed to be analyzed (No group variable! This can be achieved through `select()`), and the second is the `group` parameter specifying the group variable.

Moreover, the function `heplots::boxM()` is helpful. Instead of taking the formula to specify the parameters, using the "traditional" way to write `group = ...` works pretty well combined with pipe-friendly environment!

The drawback of the functions is that they could not search column variables automatically, and that should be specified manually. And more annoying one is that, when encountering with NAs, an error will be raised by `rstatix::box_m()`, while that's fine with `heplots::boxM()`.
```{r}
salaryData |> select(salary,salbegin) |> 
  box_m(group = jobcat)

salaryData |> select(salary,salbegin) |> 
  heplots::boxM(group = jobcat)
```

Do not forget that the assumption of homogeniety of variance for each DV is needed either! However, the traditional `levene_test()` (or its variants) can deal with one DV at a time, which is tedious when we have many DVs to check.

Hopefully, we have `heplots::leveneTests(data,group)` to deal with that issue. Combined with pipe-friendly procedures, that's fabulous!
```{r}
salaryData |>  select(salbegin,salary) |> 
  heplots::leveneTests(group = jobcat)
```

#### classical
The most annoying part is that the data has to be packed purposefully.

An alternative way to use `heplots::boxM()` has been mentioned above. We can renovate the efficiency to carry this test by replacing the "formula" specifying-pattern with the "group" one.
```{r}
heplots::boxM(cbind(salary, salbegin) ~ factor(jobcat), 
              data = salaryData)
```

## MANOVA
Use the funtion `manova()`, which is from the baseR, no special need to library packages.

Similarly, the most annoying part is that the DVs and IVs has to be packed previously.
```{r}
fit = manova(DV ~ jobcat)
```

As before, simply use `summary()` of the fitted model to see the results.
```{r}
summary(fit)  #for manova
```

## effect size
Checking the effect size is also indispensable.

Use `eta_squared()` from the package `effectsize`, which contains plenty of useful functions computing effect size under various scenarios.

Note that there's also an `eta_squared()` from the package `rstatix()`, but that can only deal with univariate ANOVA, it would fail when encountering MANOVA models.

Also note that the package `effectsize` is incorporated in `bruceR`!
```{r}
effectsize::eta_squared(fit)
```

## univariate ANOVA
Needed if MANOVA reports significant results.

`summary.aov()` will report the univariate ANOVA respectively.
```{r}
summary.aov(fit)  #for indDVidual anovas
```

## post-hoc tests
Note that the post-hoc tests here are essentially for the univariate ANOVA.

Note that the `games_howell_test()` is different from `tukey_hsd()`.

Note that for many DVs to take the post-hoc tests simultaneously, first `gather()` the targeted columns, and then take the **grouped** data for the test.
```{r comment=''}
salaryData %>% gather(key = "var", 
                      value = "value", 
                      salary, salbegin) |> 
  group_by(var) |>
  games_howell_test(value~jobcat) 
```