---
title: "Exploratory Factor Analysis"
author:
  - Zircon
date: "`r Sys.Date()`"
documentclass: ctexart
output:
  rticles::ctex:
    fig_caption: yes
    number_sections: yes
    toc: yes
    
header-includes:
  - \renewcommand{\and}{\\}
  - '\usepackage[most]{tcolorbox}'
  - '\definecolor{boxidea}{HTML}{6c8aa5}'
  - '\definecolor{boxwarn}{HTML}{b58266}'
  - '\definecolor{boxtip}{HTML}{8aa085}'
  - '\newtcolorbox{concept}[1][]{enhanced, breakable, colback=boxidea!7, colframe=boxidea, coltitle=white, fonttitle=\bfseries, title=#1, arc=2pt, left=8pt, right=8pt, top=4pt, bottom=4pt}'
  - '\newtcolorbox{pitfall}[1][]{enhanced, breakable, colback=boxwarn!7, colframe=boxwarn, coltitle=white, fonttitle=\bfseries, title=#1, arc=2pt, left=8pt, right=8pt, top=4pt, bottom=4pt}'
  - '\newtcolorbox{keypoint}[1][]{enhanced, breakable, colback=boxtip!7, colframe=boxtip, coltitle=white, fonttitle=\bfseries, title=#1, arc=2pt, left=8pt, right=8pt, top=4pt, bottom=4pt}'
  - \usepackage{amsmath}
---

# 这章在讲什么

主成分分析（PCA）和探索性因子分析（EFA）都是\textbf{降维}：把几十个高度相关的变量压缩成少数几个「主成分」/「因子」，让数据更可解释、更容易建模。两者数学很像（都用特征分解 / SVD），但**问题假设和目标不同**：

- **PCA**：纯几何变换——找一组互不相关的新坐标轴，让数据在第一根轴上方差最大、第二根次之……目标是**解释方差**。主成分本身没有「实质含义」，是数据自身的方向。
- **EFA**：假设有几个**潜在心理结构**（latent constructs）驱动了观测变量，目标是\textbf{推断这些结构是什么}。比如大五人格的 25 道题背后有 5 个潜在因子（开放性、尽责性、外向性、宜人性、神经质）。

\begin{concept}[PCA 的数学]
对中心化（标准化）后的数据 $\mathbf{X}_{n \times p}$，计算协方差矩阵 $\mathbf{S} = \frac{1}{n-1}\mathbf{X}'\mathbf{X}$，做特征分解 $\mathbf{S} = \mathbf{V}\mathbf{\Lambda}\mathbf{V}'$。

特征向量 $\mathbf{V}$ 是新坐标轴的方向，特征值 $\lambda_i$ 告诉你每根轴上有多少方差。按 $\lambda$ 从大到小排序，保留前 $k$ 个就完成了降维。

「保留多少 $k$」三种判据：(1) \textbf{Kaiser 准则} $\lambda_i > 1$（保留高于平均方差的成分）；(2) \textbf{碎石图肘部}（scree plot，看曲线突然变缓的位置）；(3) \textbf{平行分析}（parallel analysis，跟随机数据比较，更稳健，bruceR::EFA 默认就给）。
\end{concept}

\textbf{EFA 跟 PCA 的关键区别}：EFA 引入「特有方差」（unique variance）—— 每个观测变量除了被共同因子解释，还有自己独有的方差（包括测量误差）。EFA 只解释「公因子之间的协方差」，PCA 解释「所有方差」。实务上对人格 / 智力量表做心理测量分析时用 EFA；对图像 / 基因表达做降维时用 PCA。

```{r}
rm(list=ls())
library("readxl")
library("psych")
library(dplyr)
library(tidyr)
library(bruceR)
library(corrplot)
library(ggplot2)
library(GPArotation)
```


```{r}
currentData = psych::bfi
currentData = currentData %>% drop_na()
currentData = currentData[complete.cases(currentData),]
currentData = currentData[1:25]
```


```{r}
describe(currentData)
```

## Correlation Matrix

```{r}
cor_matrix = cor(currentData)
corrplot(cor_matrix, method="number")
```

## Data Adequacy

get the overall MSA from KMO; look at the bartlett test

```{r}
KMO(cor(currentData))
cortest.bartlett(cor(currentData),nrow(currentData))
det(cor_matrix)
```

## Determine num. of Factors

* To determine the number of factors, scree and parallel are two methods.
* Scree is obsolete; parallel suggests 3 factors here.

```{r}
scree(currentData)
fa.parallel(currentData,fm='pa',fa='both')
```

# Factor Analysis

## NULL Model

Take a look: without selecting the number of factors, without rotation.

Factoring Method "`fm`":
* `minres`: do a minimum residual
* `pa`: do the principal factor solution
* `ml`: do a maximum likelihood factor analysis

The Proportion Var and Cumulative Var are the variance that can be explained by factors.

"SS loadings" is the sum of squared loadings for each factor; **>1** means the factor is worth keeping.

The model can be printed with the following results:

* Standardized loadings (pattern matrix) based upon correlation matrix
  * SS loadings
    * The eigenvalues, the sum of the squared loadings for each factor. 
    * **Better to be >1**.
  * Proportion Var 
    * The overall variance (for all variables/items) that can be explained by each factor.
  * Cumulative Var
  * Proportion Explained
    * The relative importance of each factor.
  * Cumulative Proportion
* Test of the hypothesis that selected numbers of factors are sufficient
* Measures of factor score adequacy
  * Correlation of (regression) scores with factors 
  * Multiple R square of scores with factors
  * Minimum correlation of possible factor scores
* Data frame with loading matrix, communality, uniqueness, and item complexity.

**Standardized loadings (pattern matrix) based upon correlation matrix** is the most important part!

```{r}
fa_none = fa(currentData,nfactors = ncol(currentData),
             fm='minres',rotate='none')  
print(fa_none)
```

For illustration purpose:

* If using 25 factors (here only 25 items), we can see how loading and
proportion of variance decrease with factors;
* The cumulative variance explained will approach 1.

## Screened Model

According to spree plot, take 6 factors.

Note that restrictions on factor number will have practical effect on the model, instead of just trimming the model to the reduced one.

```{r}
fa6_none = fa(currentData, 6,fm='minres',rotate='none')  
print(fa6_none)
```

However, the old big five model only has 5 factors, and we're to designate this number.

```{r}
fa5_none = fa(currentData, 5,fm='minres',rotate='none')  
print(fa5_none)
```

Obviously, with more factors taken into account, more cumulative variance will be explained.

# Factor Rotation

Factor loadings listed for each variable. The square of it is the variation accounted for.

The hope is that all these correlation coefficients are close to (+/-)1 or 0, so we can easily interpret the meanings of these components (factors). We will accomplish this goal through rotation.

The relationship (correlation coefficient r) between the variable and the factor is shown. We can view it on a **Factor Plot**.

Use 2 factors for illustration purpose, to show how rotation works.

```{r}
fa_unrotated = fa(currentData,nfactors = 2,fm='minres',rotate='none')
# print(fa_unrotated)
fa_unrotated$loadings  #laodings, unrotated
```

Now, rotate it (orthogonal and oblique).

Un-rotated factors are typically not very interpretable (most factors are correlated with many variables), though it has largest variance explained.

Factors are rotated to make them more meaningful and easier to interpret (each variable is associated with a minimal number of factors). Looking for “simple structure”.

The most popular rotational method is Varimax rotations (方差最大旋转）. Varimax use orthogonal rotations yielding uncorrelated
factors/components.

Varimax attempts to maximize the variance that a single factor can account for (across all variables). This enhances the interpretability of the factors. (kind of simplification of interpretation)

```{r}
fa_otho = fa(currentData,2,fm='minres',rotate='varimax')
print(fa_otho)
```


Notice that the loading matrix changes, with values moving aside either towards 0 or 1. However, the communality doesn't change much, and this result is quite natural.

As for rotation, Oblimin is the default in R, which is an oblique rotation. Varimax is the most popular orthogonal rotation. Other rotational methods include Quartimax (Orthogonal, each variable accounted by minimal number of factors, 简化对变量的解释), Equamax (Orthogonal), Promax (oblique, this one is fast!). Oblique rotation yields correlated factors.

```{r}
fa_otho$loadings  # show the loadings after rotation
fa_otho$rot.mat  # show the rotation matrix
fa_oblique = fa(currentData,2,fm='minres',rotate='promax')
print(fa_oblique)
```

Use diagram to show how variables relate to the factors. 
```{r}
fa.diagram(fa_otho)
```

What if we use 5 factors, either rotated or not. To see whether big five really works here.

Note `fa.diagram` use r = 0.3 as a cut off, you can change it.

```{r}
fa5_none = fa(currentData,5,fm='minres',rotate='none')
fa.diagram(fa5_none)
fa5_otho = fa(currentData,5,fm='minres',rotate='varimax')
fa.diagram(fa5_otho)
```


Interestingly, the 6-factor model does not work that well.

```{r}
fa6_none = fa(currentData,6,fm='minres',rotate='none')
fa.diagram(fa6_none)
fa6_otho = fa(currentData,6,fm='minres',rotate='varimax')
fa.diagram(fa6_otho)
```

Scores, or the coefficient for the model: Factors = scores*X where X are the original data, Factors are the obtained latent variables, scores are their coefficients.

The scores here are for normalized data, which is the case for SPSS.

```{r}
head(fa_otho$scores)
```

The below is for you to hand check the meaning of SS laoding, uniqueness and communality.

Communality, the variance that can be explained by factors, which should be high if the model is good.

```{r}
# communality
apply(fa_otho$loadings^2,1,sum)
# uniqueness
1 - apply(fa_otho$loadings^2,1,sum)
```








```{r}
P = EFA(data = currentData,vars = names(currentData))
P$eigenvalues
P$loadings
P$result
P$result.kaiser
P$scree.plot
```
```{r}

```



# Appendix

## `psych::describe()`

# 容易踩的坑

\begin{pitfall}[PCA / EFA 最常见的几个误区]
\begin{enumerate}
\item \textbf{不标准化就跑 PCA}：与聚类同理——方差大的变量会主导第一主成分。除非所有变量天然同量纲（如百分比、Z 分数），都应该先 \texttt{scale()} 或者用相关矩阵代替协方差矩阵跑 PCA（\texttt{prcomp(x, scale. = TRUE)}）。
\item \textbf{把 PCA 当 EFA 用}：用 \texttt{princomp} 或 \texttt{prcomp} 算出的「载荷」叫主成分载荷，本质是特征向量；EFA 的载荷是因子载荷，会随旋转方法变化。两者经过同一份数据可能数值很接近，但解释起来\textbf{完全不同}：主成分是「数据 → 成分」的投影，因子是「因子 → 数据」的反向解释。
\item \textbf{Kaiser 准则保留所有 $\lambda > 1$ 成分}：Kaiser ($\lambda > 1$) 经常会保留过多成分，特别是变量数 $p > 30$ 时。优先信赖\textbf{平行分析}（parallel analysis）和\textbf{碎石图}的肘部，Kaiser 当作下限参考。
\item \textbf{斜交旋转后看模式矩阵 / 结构矩阵搞混}：斜交旋转（如 promax / oblimin）后会产生两个载荷矩阵：\textbf{模式矩阵}（pattern matrix，回归系数）报告每个变量在每个因子上的「独特贡献」；\textbf{结构矩阵}（structure matrix，相关系数）报告每个变量和因子的「总相关」。解释因子用模式矩阵（剔除因子间相关之后的纯净贡献），别看错。正交旋转（varimax）两个矩阵相同。
\end{enumerate}
\end{pitfall}
