\documentclass[oneside]{book}

\usepackage{amsmath, amsthm, amssymb, amsfonts}
\usepackage{thmtools}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{geometry}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{framed}
\usepackage[dvipsnames]{xcolor}
\usepackage{environ}
\usepackage{tcolorbox}
\tcbuselibrary{theorems,skins,breakable}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{multicol}
\usetikzlibrary{positioning, arrows.meta, patterns}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{cancel}
\usepackage{booktabs}
\usepackage{enumitem}
\setlist[itemize]{leftmargin=*,labelindent=0pt}

\usepackage{fancyhdr}
\fancypagestyle{plain}{\fancyhf{}\fancyfoot[C]{\thepage}}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{\textit{\uppercase{\leftmark}}}
\fancyhead[R]{\thepage}
\fancyfoot[R]{\tiny Intermediate Econometrics \textemdash\ ruizhou03.com}

\setstretch{1.2}
\geometry{
    textheight=9in, textwidth=5.5in, top=1in,
    headheight=12pt, headsep=25pt, footskip=30pt
}

\def\notetitle{Intermediate Econometrics}
\def\notesubtitle{A Self-Contained Course Companion}
\def\noteauthor{Lecture course by \textbf{Prof. Xiaojun Song} \\
    Guanghua School of Management, Peking University \\
    \vspace{0.5em}
    Notes written and {\LaTeX}-typeset by Rui Zhou}
\def\notedate{Spring 2023 \\ \small (rewritten and unified, 2026)}

\input{theorems.tex}
\input{commands.tex}

\begin{document}

\title{\textbf{
    \LARGE{\notetitle}\\
    \vspace{0.5em}
    \large\textit{\notesubtitle}
    \vspace*{8\baselineskip}}
}
\author{\noteauthor}
\date{\notedate}

\maketitle
\newpage

\thispagestyle{empty}
\section*{How to Read This Book}

This is a course companion for a one-semester \textbf{intermediate econometrics} course based on Wooldridge's \emph{Introductory Econometrics}. It is written to be read straight through, like a short textbook, rather than skimmed like a formula sheet.

The first half of the course covers the mechanics of the linear regression model: how ordinary least squares (OLS) is defined, when it is unbiased, what its variance is, and how to test hypotheses about its coefficients. The second half relaxes the convenient assumptions one at a time --- nonlinear functional forms, qualitative regressors, heteroskedasticity, problematic data, panel structure, and endogeneity --- and shows what tool repairs each broken assumption. Every later chapter is, at heart, a story of the form ``\emph{assumption~X fails; here is what goes wrong, how we detect it, and how we fix it.}''

\subsection*{How Each Chapter Is Organized}

Each chapter opens with a short \emph{story} in plain English: what problem are we solving, and why are we about to introduce a particular tool? We then formalize the idea, state the main results, and work through the algebra. Definitions, theorems, and assumptions are boxed so you can find them quickly when reviewing.

\subsection*{Box Color Code}

\begin{itemize}
    \item \tcbox[colback=green!10!white, colframe=green!10!white, size=fbox]{\textbf{Green: Definition}} --- the objects and concepts we build on.
    \item \tcbox[colback=cyan!10!white, colframe=cyan!10!white, size=fbox]{\textbf{Blue: Theorem}} --- the main results, usually with a proof or sketch.
    \item \tcbox[colback=pink!20!white, colframe=pink!20!white, size=fbox]{\textbf{Pink: Assumption}} --- the conditions we impose (e.g.\ SLR.1--SLR.5); we cite them by name.
    \item \tcbox[colback=violet!10!white, colframe=violet!10!white, size=fbox]{\textbf{Purple: Lemma}} --- intermediate results used in a proof.
    \item Worked \textbf{examples} and \textbf{remarks} appear in lightly ruled boxes throughout.
\end{itemize}

\subsection*{A Note on Notation}

Throughout, $n$ is the sample size, $k$ the number of regressors (excluding the intercept), $u$ the population error, and $\widehat{u}_i$ the OLS residual. We write $\E{\cdot}$ for expectation, $\Var{\cdot}$ for variance, $\Cov{\cdot}{\cdot}$ for covariance, and $\plim$ for the probability limit. A hat denotes an estimator ($\hb_j$); a tilde denotes an estimator from a \emph{misspecified} (e.g.\ short) regression ($\widetilde{\beta}_j$).

\newpage

\setcounter{tocdepth}{1}
\tableofcontents
\newpage

\include{chapters/ch1_simple_regression}
\include{chapters/ch2_mlr_estimation}
\include{chapters/ch3_mlr_inference}
\include{chapters/ch4_asymptotics}
\include{chapters/ch5_further_issues}
\include{chapters/ch6_dummies}
\include{chapters/ch7_heteroskedasticity}
\include{chapters/ch8_specification}
\include{chapters/ch9_panel}
\include{chapters/ch10_iv_2sls}

\end{document}
